---
title: 'Notes on Ch2: Geographic data in R'
author: "Norman LIm"
date: "2025-06-28"
output:
  pdf_document: default
  html_document: default
---

## Introduction

Fundamental geographic data models:
- vector: represents the world using points, lines, and polygons  
- raster: divides the surface up into cells of constant size. Example: background images used in web maps  

Which one to use? It depends:
- for social science, vector data tends to dominate because human settlements tend to have discrete borders  
- for environmental sciences, raster dominates because of reliance on sensor data  


## Vector data
-based on points located within a coordinate reference system (CRS)  

important low-level libraries in the sf package include:  
- GDAL  
- PROJ  
- GEOS: a planar (aka flat or projected) geometry engine  
- S2: a spherical (unprojected) geometry engine  


### Simple features
- simple features is an open standard developed by the Open Geospatial Consortium (OGC)  
- there are 18 geometry types supported by the simple features specification  
- only 7 are used in most geographic research  
- these 7 are fully supported in R via the **sf** package  

Simple features fully supported in R via **sf**:  
- multilinestring  
- multipoint  
- multipolygon  
- geometrycollection  
- polygon 
- point 
- linestring 

load libraries:
```{r}
library(sf)          # classes & functions for vector data
library(terra)       # classes & functions for raster data
library(spData)      # for geographic data
library(spDataLarge) # for larger geographic data
```


In sf,  
- simple feature objects in R are stored in a data frame  
- geographic data are usually contained in `geom` or `geometry` column  


```{r}
class(world)
```

```{r}
names(world)
```

- in the `world` data, the `geom` column contains a list of all the coordinates of the country polygons 

- `sf` objects can be plotted quickly using the function `plot()` or `plot.sf()`  

```{r}
plot(world)
```


Here is where R shines --- making statistical calculations --- on "geospatial data"  

```{r}
summary(world["lifeExp"])
```

Note: if we use the `$` to access the column `lifeExp`, we won't see the geom:  
```{r}
summary(world$lifeExp)
```

Subsetting `sf` objects
```{r}
world_mini = world[1:2, 1:3]
world_mini
```

Note: again, we use more pythonic syntax here.

Trying the more r-esque sytax:  
```{r}
world_mini2 <- world[1:2, 1:3]
world_mini2
```

- the authors of the book prefer using the `equals assignment` notation for faster typing and compatibility with other languages such as Python and Javascript.  


### Why simple features?
- faster reading and writing of data  
- enhanced plotting performance  
- `sf` objects can be treated as data frames in most operations  
- consistent and intuitive  
- works with the pipe operator  


Importing geographic vector data
```{r}
world_dfr = st_read(system.file("shapes/world.gpkg", package = "spData"))
```

```{r}
world_tbl = read_sf(system.file("shapes/world.gpkg", package = "spData"))

class(world_dfr)
class(world_tbl)
```


## Basic maps
```{r}
plot(world[3:6])
plot(world["pop"])
```

- plots can be added as layers to existing images by setting `add = TRUE`  

for example:
```{r}
world_asia = world[world$continent == "Asia",]
asia = st_union(world_asia)
```

Plotting Asia on top of the world map as layer:
```{r}
plot(world["pop"], reset = FALSE)
plot(asia, add = TRUE, col = "red")
```

Overlaying circles on plot of map:
```{r}
plot(world["continent"], reset = FALSE)
cex = sqrt(world$pop) / 10000
world_cents = st_centroid(world, of_largest = TRUE)
plot(st_geometry(world_cents), add = TRUE, cex = cex)
```

- in the code, the function `st_centroid()` was used to "convert" polygons to points. The aesthetics are controlled by the `cex` argument.  


Plotting with Expanded bounding box
```{r}
india = world[world$name_long == "India", ]
plot(st_geometry(india), expandBB = c(0, 0.2, 0.1, 1), col= "gray", lwd = 3)
plot(st_geometry(world_asia), add = TRUE)
```


### Geometry types

Standard encoding for simple feature geometries:  
- well-known binary (WKB): usually in hexadecimal  
- well-known text (WKT): human-readable text  


Commonly used geometry types in `sf`:
- point: coordinates in two-, three-, or four-dimensional space. Example: `POINT (5 2)`  
- linestring: a sequence of points with a straight line connecting the points. Example: `LINESTRING (1 5, 4 4, 4 1, 2 2, 3 2)`  
- polygon: a sequence of points that form a closed, non-intersecting ring. Example: `POLYGON ((1 5, 2 2, 4 1, 4 4, 1 5))`  
- Multipoint: Example: `MULTIPOINT (5 2, 1 3, 3 4, 3 2)`  
- Multilinestring: Example: `MULTILINESTRING ((1 5, 4 4, 4 1, 2 2, 3 2), (1 2, 2 4))`  
- Multipolygon: Example: `MULTIPOLYGON (((1 5, 2 2, 4 1, 4 4, 1 5), (0 2, 1 2, 1 3, 0 3, 0 2)))`  
- Geometry collection: any combination of the geometries above. Example: `GEOMETRYCOLLECTION (MULTIPOINT (5 2, 1 3, 3 4, 3 2), LINESTRING (1 5, 4 4, 4 1, 2 2, 3 2))`  


Combining geometries and non-geographic attributes
```{r}
lnd_point = st_point(c(0.1, 51.1))
lnd_geom = st_sfc(lnd_point, crs = "ESPG:4326")
lnd_attrib = data.frame(
  name = "London",
  temperature = 25,
  date = as.Date("2023-06-21")
)
lnd_sf = st_sf(lnd_attrib, geometry = lnd_geom)

lnd_sf
```

- if we need to "make"draw" geometries from scatch, we add an `st_` prefix to the geometry type listed above  

```{r}
st_point(c(5, 2))
st_point(c(5, 2, 3))
st_point(c(5, 2, 1), dim = "XYM")
st_point(c(5, 2, 3, 1))
```

- XY is 2d, XYZ is 3d, XYZM is 3d with additional variable (like for improving measurement accuracy)  

```{r}
# multipoint
multipoint_matrix = rbind(c(5, 2), c(1, 3), c(3, 4), c(3, 2))
st_multipoint(multipoint_matrix)

# linestring
linestring_matrix = rbind(c(1, 5), c(4, 4), c(4, 1), c(2, 2), c(3, 2))
st_linestring(linestring_matrix)
```

Multi-sf examples:
```{r}
## POLYGON
polygon_list = list(rbind(c(1, 5), c(2, 2), c(4, 1), c(4, 4), c(1, 5)))
st_polygon(polygon_list)

## POLYGON with a hole
polygon_border = rbind(c(1, 5), c(2, 2), c(4, 1), c(4, 4), c(1, 5))
polygon_hole = rbind(c(2, 4), c(3, 4), c(3, 3), c(2, 3), c(2, 4))
polygon_with_hole_list = list(polygon_border, polygon_hole)
st_polygon(polygon_with_hole_list)

## MULTILINESTRING
multilinestring_list = list(rbind(c(1, 5), c(4, 4), c(4, 1), c(2, 2), c(3, 2)), 
                            rbind(c(1, 2), c(2, 4)))
st_multilinestring(multilinestring_list)

## MULTIPOLYGON
multipolygon_list = list(list(rbind(c(1, 5), c(2, 2), c(4, 1), c(4, 4), c(1, 5))),
                         list(rbind(c(0, 2), c(1, 2), c(1, 3), c(0, 3), c(0, 2))))
st_multipolygon(multipolygon_list)

## GEOMETRYCOLLECTION
geometrycollection_list = list(st_multipoint(multipoint_matrix),
                              st_linestring(linestring_matrix))
st_geometrycollection(geometrycollection_list)
```


### Simple feature columns (sfc)
- a list of `sfg` objects, where an `sfg` object contains only a single simple feature geometry.  
- to combine simple features in a objet with two features, we can use the `st_sfc()`  
- this looks like a function analogous to the concatenate function in r-base (`c()`):  
```{r}
point1 = st_point(c(5, 2))
point2 = st_point(c(1, 3))

points_sfc = st_sfc(point1, point2)
points_sfc
```

- when `sfc` objects contain objects of the same geometry type, they can be converted into an `sfc` of polygon type  
- the geometry type of an object can be verified using `st_geometry_type()`:  
```{r}
polygon_list1 = list(rbind(c(1, 5), c(2, 2), c(4, 1), c(4, 4), c(1, 5)))
polygon1 = st_polygon(polygon_list1)

polygon_list2 = list(rbind(c(0, 2), c(1, 2), c(1, 3), c(0, 3), c(0, 2)))
polygon2 = st_polygon(polygon_list2)

polygon_sfc = st_sfc(polygon1, polygon2)
st_geometry_type(polygon_sfc)
```

- multilinesgring examples:
```{r}
# sfc MULTILINESTRING
multilinestring_list1 = list(rbind(c(1, 5), c(4, 4), c(4, 1), c(2, 2), c(3, 2)), 
                            rbind(c(1, 2), c(2, 4)))
multilinestring1 = st_multilinestring((multilinestring_list1))
multilinestring_list2 = list(rbind(c(2, 9), c(7, 9), c(5, 6), c(4, 7), c(2, 7)), 
                            rbind(c(1, 7), c(3, 8)))
multilinestring2 = st_multilinestring((multilinestring_list2))
multilinestring_sfc = st_sfc(multilinestring1, multilinestring2)
st_geometry_type(multilinestring_sfc)
```


- it is also possible to create an `sfc` object from `sfg` objects with different geometry types:  
```{r}
point_multilinestring_sfc = st_sfc(point1, multilinestring1)
st_geometry_type(point_multilinestring_sfc)
```

- `sfc` objects can store additional information on the coordinate reference system (CRS).  
- the crs information can be verified with `st_crs()`:  
```{r}
st_crs(points_sfc)
```

- all geometries in `sfc` objects must have the same CRS. 
- the functions `st_sfc()` or `st_sf()` can be used to specify the CRS. 
```{r}
# Set the crs with an identifier referring to an "EPSG" CRS code
points_sfc_wgs = st_sfc(point1, point2, crs = "EPSG:4326")
st_crs(points_sfc_wgs)
```


### The `sfheaders` package
- speeds up the construction, conversion, and manipulation of `sf` objects.  

Example: a vector converted to `sfg_POINT`
```{r}
v = c(1, 1)
v_sfg_sfh = sfheaders::sfg_point(obj = v)
v_sfg_sfh
```

Example: creation of `sfg` objects from matrices and data frames:
```{r}
# from matrix
m = matrix(1:8, ncol = 2)
sfheaders::sfg_linestring(obj = m)

# from data frame
df = data.frame(x = 1:4, y = 4:1)
sfheaders::sfg_polygon(obj = df)
```

Other examples:
```{r}
sfheaders::sfc_point(obj = v)
sfheaders::sfc_linestring(obj = m)
sfheaders::sfc_polygon(obj = df)

sfheaders::sf_point(obj = v)
sfheaders::sf_linestring(obj = m)
sfheaders::sf_polygon(obj = df)
```


Example: defining the CRS on an object created by using `sfheaders`:  
```{r}
df_sf = sfheaders::sf_polygon(obj = df)
st_crs(df_sf) = "EPSG:4326"
```


### Spherical geometry operations with S2
- the S2 geometry engine is turned on by default   

To verify if S2 engine is set to on or off:  
```{r}
sf_use_s2()
```


What if the S2 engine is turned off?
```{r}
india_buffer_with_s2 = st_buffer(india, 1) # 1 meter
sf_use_s2(FALSE)

india_buffer_without_s2 = st_buffer(india, 1) # 1 degree
```

Leaving the S2 engine turned on, unless explicitly stated:
```{r}
sf_use_s2(TRUE)
```

### Raster data
- represents the world with continuous grid of cells aka pixels  
- usually consists of a raster header and a matrix that represent the pixels  
- the raster header defines the CRS, origin, and extent  
- the origin is usually the lower left corner    
- in `terra`, the origin is the upper left corner  

Resolution formula:
$$
\text{resolution} = \frac{\text{xmax}-\text{xmin}}{\text{ncol}}, \frac{\text{ymax}-\text{ymin}}{\text{nrow}}
$$

We can easily access and modify each single cell by:
- using the ID of a cell  
- explicitly specifying the rows and columns  


### R packages for working with raster data
- **terra**  
- **stars**  

Converting **terra** objects to **stars**:
- `st_as_stars()`

Converting **stars** objects to **terra**:
- `rast()`


### Introduction to **terra**
- terra provides the possibility to divide the raster into smaller chunks and process them iteratively instead of loading the whole raster file into RAM  

Example: basic terra usage

Creation of a `SpatRaster` object
```{r}
raster_filepath = system.file("raster/srtm.tif", package = "spDataLarge")
my_rast = rast(raster_filepath)
class(my_rast)
```

Inspection of the raster header:
```{r}
my_rast
```

Using dedicated functions to see a report on each raster component:
```{r}
dim(my_rast)
ncell(my_rast)
res(my_rast)
ext(my_rast)
crs(my_rast)
inMemory(my_rast)
```

### Basic map-making

Plotting the raster file
```{r}
plot(my_rast)
```


### Raster classes

- The `SpatRaster` class represents rasters object of **terra**
```{r}
single_raster_file = system.file("raster/srtm.tif", package = "spDataLarge")
single_rast = rast(raster_filepath)
```

Creating new rasters from scratch with `rast()`
```{r}
new_raster = rast(
  nrows = 6, ncols = 6, 
  xmin = -1.5, xmax = 1.5, 
  ymin = -1.5, ymax = 1.5,
  vals = 1:36
)
```

- based on the resolution formula above, the example has a resolution of 0.5 degrees.  
- the unit is degrees since the default CRS of raster objects is WGS84.  
- the CRS can be specified using the `crs` argument    

The `SpatRaster` class also handles multiple layers:
```{r}
multi_raster_file = system.file("raster/landsat.tif", package = "spDataLarge")
multi_rast = rast(multi_raster_file)
multi_rast
```


Displaying the number of layers with `nlyr()`:
```{r}
nlyr(multi_rast)
```

- in multi-layer raster objects, layers can be "subset" using `[[]]` or `$`  
- `terra::subset()` can also be used to select layers  
```{r}
multi_rast3 = subset(multi_rast, 3)
multi_rast4 = subset(multi_rast, 4)
```


- just like in r-base, combining can be accomplished using the comcatenate operator:
```{r}
multi_rast34 = c(multi_rast3, multi_rast4)
```


### Geographic coordinate reference systems
- longitude: East-West angular distance from the Prime Meridian plane    
- latitude: North-South angular distance from the equatorial plane  
- Earth is assumed to be spherical (less accurate) or ellipsoidal (more accurate)  
- equatorial radius is about 11.5 km longer than the polar radius  
- for the ellipsoidal model, what ellipsoid to use is defined by the *datum*  
- geocentric datum (WGS84) has its center located in the Earth's center of gravity  
- local datum (e.g. NAD83) the ellipsoidal surface is shifted to align with the surface at a particular location  
- with  local datum, local variations in the earth's surface are accounted for  


### Projected coordinate reference systems
- the three-dimensional surface of the Earth is "projected" on a flat surface    
- the transformation introduces some deformations  
- some parts get distorted  
- main projection types include conic, cylindrical, and planar  


### Units
- most CRSs use meters, but some use feet  
```{r}
luxembourg = world[world$name_long == "Luxembourg",]
st_area(luxembourg)
```

- be careful with the units when doing calculations  
- if we simply divide the area with the conversion factor 1000000 to convert the area to sq. km:
```{r}
st_area(luxembourg) / 1000000
```
the result is still in square meters. This is not correct.  

- the better way is to use the **units** package to convert units:
```{r}
units::set_units(st_area(luxembourg), km^2)
```


- the `res()` command will not show the units of the raster data  
- only **sf** supports units  
- we have to know the units used in the projection of choice  


## Exercises
E1. Use `summary()` on the geometry column of the `world` data object that is included in the **spData** package. What does the output tell us about:

    Its geometry type?
    The number of countries?
    Its coordinate reference system (CRS)?

Solution:
```{r}
summary(world$geom)
```

Ans: Multipolygon, 177 countries, ESPG:4326

E2. Run the code that ‘generated’ the map of the world in Section 2.2.3 (Basic map-making). Find two similarities and two differences between the image on your computer and that in the book.

    What does the `cex` argument do (see `?plot`)?
    Why was `cex` set to the `sqrt(world$pop) / 10000`?
    Bonus: experiment with different ways to visualize the global population.

Solution:
```{r}
plot(world["continent"], reset = FALSE)
cex = sqrt(world$pop) / 10000
world_cents = st_centroid(world, of_largest = TRUE)
plot(st_geometry(world_cents), add = TRUE, cex = cex)
```

Ans: 
Two similarities: 
1. map color.  
2. circle marks.   

Two differences:
1. legend
2. title

`cex` controls the symbol size. `cex` is set to the square root of population in order to make the area of the circle proportional to the population -- doubling the population will correspond to a circle with twice the area. If we don't use sqrt, then we will effectively double the diameter instead.


E3. Use `plot()` to create maps of Nigeria in context (see Section 2.2.3).

    Adjust the `lwd`, `col` and `expandBB` arguments of plot().
    Challenge: read the documentation of `text()` and annotate the map.

Solution:
```{r}
world_africa = world[world$continent == "Africa", ]
plot(st_geometry(world_africa))

nigeria = world[world$name_long == "Nigeria", ]
plot(
  st_geometry(nigeria), 
  expandBB = c(0, 0.1, 0.2, 2), 
  col = "gray", 
  lwd = 4
)

# world_africa = world[world$continent == "Africa", ]
plot(st_geometry(world_africa), add = TRUE)
text(7.985654, 9.544975, "Nigeria")
```


E4. Create an empty `SpatRaster` object called `my_raster` with 10 columns and 10 rows. Assign random values between 0 and 10 to the new raster and plot it.

Solution:
```{r}
# x11()
# plot(my_raster, 
#     main = "My Raster", 
#     breaks = c(50, 100, 150, 200),
#     mar = c(4, 4, 2, 2)
# )
```


E5. Read-in the `raster/nlcd.tif` file from the **spDataLarge** package. What kind of information can you get about the properties of this file?

Solution:
```{r}
raster_file = system.file("raster/nlcd.tif", package = "spDataLarge")
exer5_rast = rast(raster_file)
exer5_rast
```

Ans:
Class is SpatRaster. Size is 1359 x 1073 (cols, rows). Number of layers: 1. Resolution (31.5, 31.5).
xmin = 301903.3, xmax = 335735.4, 
ymin = 4111244, ymax = 4154086, 
CRS = ESPG:26912, datum = NAD83
and others.

E6. Check the CRS of the `raster/nlcd.tif` file from the **spDataLarge** package. What kind of information you can learn from it?

Soln:
Datum is NAD83 or the North American Datum of 1983.
Coordinate Reference System is ESPG:26912. This is a projected coordinate system common in North America.






